\bf Estimating the number of clusters using cross-validation

### 2016-06-14 cvclust.tex
revise through sec 4.1

A clustering procedure segments a collection of items into smaller groups,
with the property that items in the same group are more similar to each other
than items in different groups \citep{hartigan1975clustering}.  These
procedures are used in two main applicaitons: (a) exploratory analysis, where clusters
reveal homogeneous sub-groups within a large sample \cite{TODO}; (b) data
reduction, where high-dimensional item attribute vectors get reduced discrete
cluster labels \citep{jain1999data}.

@@

Several authors advocate choosing $k$ by performing a sequence of hypothesis
tests will null and alternative hypotheses of the form $H_0 : k = k_0$ and
$H_1: k > k_0$, starting with $k_0 = 1$ and proceeding sequentially with
higher values of $k_0$ until a test fails to reject $H_0$. The gap statistic
method typifies this class of methods, with a test statistic that measures the
within-cluster dispersion relative to what is expected under a reference
distribution \citep{tibshirani2001estimating}. Similar methods include those
developed by \citet{TODO}.

Other authors have proposed choosing $k$ by using information
criteria.  \citet{sugar2003finding} proposed an approach that
minimizes the estimated ``distortion'', the average distance per dimension.
\citet{fraley2002model}'s model-based method fits Gaussian mixture model
models to the data, then selects the number of mixture components, $k$, using
the Bayesian Information Criterion (BIC).

@@

The procedure we propose in this report is based on a form of
cross-validation, and it is adaptive to the characteristics of the data
distribution. The essential idea is to devise a way to measure a form 
of internal prediction error associated each choose of $k$, and then choose
the $k$ with the smallest associated error.

In this report, we extend the Wold and Gabriel methods to the clustering
problem, specifically to choose an appropriate number of clusters for a
dataset.  We prove that the Gabriel method is self-consistent, and we analyze
some of its properties in the presence of noise.  We compare these methods to
state-of-the-art algorithms, and show that both are competitive.

\textit{TODO: Add outline}

\section{Cross-validation for selecting the number of clusters}
\label{sec:meth}
Cross-validation is commonly used for model selection in supervised learning
problems.  In these settings, the data comes in the form of $N$
predictor-response pairs, $(X_1, Y_1), \dotsc, (X_N, Y_N)$, with $X_i \in
\R^{p}$ and $Y_i \in \R^{q}$.  The data can be represented as a matrix with
$N$ rows and $p + q$ columns.  We partition the data into $K$ hold-out
``test'' subsets, with $K$ typically chosen to be $5$ or $10$.  For each
``fold'' $r$ in the range $1, \dotsc, K$, we permute the rows of the data
matrix to get $\dataX$, a matrix with the $r$th test subset as its trailing
rows.  We partition $\dataX$ as
\[
  \dataX =
  \begin{bmatrix}
    \Xtrain & \Ytrain \\
    \Xtest  & \Ytest
  \end{bmatrix}.
\]
We use the training rows $[ \Xtrain\ \Ytrain ]$ to fit a regression model
$\hat Y = \hat Y(X)$, and then evaluate the performance of this model on the
test set, computing the cross-validation error $\|\Ytest - \hat Y(\Xtest)\|^2$
or some variant thereof.  We choose the model with the smallest
cross-validation error, averaged over all $K$ folds.

In unsupervised learning problems like factor analysis and clustering, the
features of the observations are not naturally partitioned into ``predictors''
and ``responses'', so we cannot directly apply the cross-validation procedure
described above.  For factor analysis, there are at least two versions of
cross-validation.  \citet{wold78cross} proposed a ``speckled'' holdout, where
in each fold we leave out a subset of the elements of the data matrix.  Wold's
procedure works well empirically, but does not have any theoretical support,
and it requires a factor analysis procedure that can handle missing data.
\citet{owen2009bi} proposed a scheme called ``bi-cross-validation'' wherein
each fold designates a subset of the data matrix columns to be response and a
subset of the rows to be test data.  This generalized a procedure due to
\citet{gabriel2002biblot}, who proposed holding out a single column and a
single row at each fold.  Owen and Perry proved that this procedure is
self-consistent, in the sense that it performs the correct model selection in
the absence of noise, and \citet{perry2009cross} provided more theoretical
support.

@@

We now give the details of how to implement
the Gabriel cross-validation to locate the optimal cluster number $k$. The
Wold cross-validation algorithm is described in Appendix $A$.

\subsection{Gabriel CV algorithm}
\label{sec:gabriel-cv-algorithm}
We are given a data matrix with $N$ rows and $P$ columns.  In each fold of
cross-validation, we permute the rows and columns of the data matrix and then
partition the rows and columns as $N = n + m$ and $P = p + q$ for 
non-negative integers $n$, $m$, $p$, and $q$.  We treat the first $p$
columns as ``predictors'' and the last $q$ columns as ``responses'';
similarly, we treat the first $n$ rows as ``training'' and the last $m$ rows
as ``test''.  In block form, the permuted data matrix is
\[
  \dataX
  =
  \begin{bmatrix}
    \Xtrain & \Ytrain \\
    \Xtest  & \Ytest
  \end{bmatrix},
\]
where
$\Xtrain \in \R^{n \times p}$,
$\Ytrain \in \R^{n \times q}$,
$\Xtest \in  \R^{m \times p}$,
and
$\Ytest \in  \R^{m \times q}$.

@@

\section{Self-Consistency of Gabriel CV method}

This section gives the self-consistency proof of the proposed Gabriel method.
Specifically, we will show that under appropriate conditions, in the absence
of noise, the Gabriel cross-validation procedure finds the optimal number of
clusters.

Because $k$-means algorithm is essential to the method, we review the
procedure here.  Given a set of observations $\{ x_1, \dotsc ,x_n \}$, and a
specified the number of clusters $k$, the goal of the $k$-means procedure is
to find a set of $k$ or cluster centers $A = \{ a_1, \dotsc, a_k \} $
minimizing the within cluster dispersion
\[
  W(A) = \sum_{i=1}^{n} \min_{a \in A} \|x_i - a\|^2.
\]
This implicitly defines a cluster assignment rule
\[
  g(x) = \argmin_{g \in \{1, \dotsc, k\}} \|x - a_g\|^2,
\]
with ties broken arbitrarily.  We will assume that the $k$-means procedure
finds an optimal solution, $A$, but we will not assume that this solution is
unique.

It will suffice to analyze a single fold of the cross-validation procedure.
As in in section~\ref{sec:gabriel-cv-algorithm} we assume that the $P$
variables of the data set have been partitioned into $p$ predictor variables
represented in vector~$X$ and $q$ response variables represented in
vector~$Y$.  The $N$ observations have been divided into two sets: $n$ train
observations and $m$ test observations.  The following theorem gives
conditions for Gabriel CV to recover the true number of clusters in the
absence of noise.

@@

This theorem is implied by the following two lemmas.

\begin{lemma}
Suppose that the assumptions of Theorem~\ref{thm:self-consistency} are in
force.  If $k < K$, then $\CV(k) > 0$.
\end{lemma}
\begin{proof}
By definition,
\[
  \CV(k)
    =
      \sum_{i=n+1}^{n+m}
        \| Y_i - \bmuY (\hGX_i) \|^2,
\]
where $\bmuY(g)$ is the center of cluster $g$ returned from applying $k$-means
to $Y_1, \dotsc, Y_n$.  Assumptions~\ref{asn:self-consistency-noiseless}
and~\ref{asn:self-consistency-test}, imply that as $i$ ranges over the test
set $n+1, \dotsc, n+m$, the response $Y_i$ ranges over all distinct values in
$\{ \muY(1), \dotsc, \muY(K) \}$.
Assumption~\ref{asn:self-consistency-distinct-muy} implies that there are
exactly $K$ such distinct values.  However, there are only $k$ distinct values
of $\bmuY(g)$.  Thus, at least one summand
\(
  \| Y_i - \bmuY(\hGX_i) \|^2
\)
is nonzero.  Therefore,
\(
  \CV(k) > 0.
\)
\end{proof}

\begin{lemma}
Suppose that the assumptions of Theorem~\ref{thm:self-consistency} are in
force.  If $k \geq K$, then $\CV(k) = 0$.
\end{lemma}
\begin{proof}
From assumptions~\ref{asn:self-consistency-noiseless},
\ref{asn:self-consistency-distinct-muy},
and~\ref{asn:self-consistency-train}, we know the cluster centers
gotten from applying $k$-means to $Y_1, \dotsc, Y_n$ must include
$\muY(1), \dotsc, \muY(K)$.  Without loss of generality, suppose that
$\bmuY(g) = \muY(g)$ for $g = 1, \dotsc, K$.  This implies that
$\hGY_i = G_i$ for $i = 1, \dotsc, n$.  Thus, employing
assumption~\ref{asn:self-consistency-noiseless} again, we get that
$\bmuX(g) = \muX(g)$ for $g = 1, \dotsc, K$.

@@

\section{Analysis under Gaussian noise}

@@

Theorem~\ref{thm:self-consistency} tells us that the Gabriel cross-validation
method recovers the true number of clusters then the noise is negligible.
While this result gives us some assurance that the procedure is well-behaved,
we can bolster our confidence and gain insight into its workings by analyzing
its behavior in the presence of noise.  We first study the case of a single
cluster in two dimensions with correlated Gaussian noise.

@@

We confirm this result with a simulation.  We perform $10$ replicates.  In each
replicate, we generate $20000$ observations from a mean-zero bivariate normal
distribution with unit marginal variances and correlation $\rho$.  We perform
a single $2 \times 2$ fold of Gabriel cross-validation and report the
cross-validation mean squared error for the number of clusters $k$ ranging
from $1$ to $5$.  Figure~\ref{fig:nullcorr-equal} shows the cross-validation
errors for all $10$ replicates.  The simulation demonstrates that in the
Gabriel cross-validation criterion chooses the correct answer $k = 1$ whenever
$\rho < 0.5$; the criterion chooses $k = 2$ clusters whenever $|\rho| > 0.5$.

\begin{figure}[H]
\centering
\includegraphics[width=3in]{demo/nullcorr/equal.pdf}
\caption{Cross-validation error on $10$ replicates, with the number of
clusters $k$ ranging from $1$ to $5$.  Data is generated from two-dimensional
multivariate normal distribution with correlation $\rho$.  The Gabriel
cross-validation criterion chooses the correct answer $k = 1$ whenever
$\rho < 0.5$; the criterion chooses $k = 2$ clusters whenever $|\rho| > 0.5$.}
\label{fig:nullcorr-equal}
\end{figure}

The reason why Gabriel CV method tends to select larger $k$ when correlation is high 
between dimensions is that it resembles the naive method mentioned in section \ref{sec:intro} 
under such circumstance. In fact, if $X$, $Y$ are perfectly correlated, Gabriel method is 
equivalent to the naive method.

@@

\textit{todo: add and revise old proof}

@@

\section{Adjusting for correlation}

\textit{TODO: add more discussion here}

When the correlation between dimensions are high, the proposed Gabriel CV
method tends to overestimate the number $k$. A simply remedy for the high
correlation is available if we assume common covariance structure among the
$k$ clusters.

\begin{enumerate}
	\item Apply Gabriel CV method on the original data $\dataX$, get estimated number of cluster $\hat{k}$
	\item Estimate the pooled covariance matrix $\hat{\Sigma}$ from the $\hat{k}$ clusters.
	\item Let $\hat{\Sigma} = \Gamma\Lambda\Gamma'$ be the Eigendecomposition of $\hat{\Sigma}$, we rescale and rotate the original data $\dataX$ to get $ \widetilde{\dataX} = \dataX\Gamma\Lambda^{-1/2}Q$, where $Q$ is a random  orthonormal rotation matrix. 
	\item Apply Gabriel CV method again on the transformed data $ \widetilde{\dataX}$ to estimate $k$. 
\end{enumerate}

\section{Performance in simulations}

@@

\section{Empirical validation}

@@

\section{Application}

\textit{TODO: add data applicaiton where number of clusters is unknown}

\section{Discussion}
\label{sec:conc}

>>>>

A clustering procedure segments a collection of items into smaller groups,
with the property that items in the same group are more similar to each other
than items in different groups \citep{hartigan1975clustering}.  Such
procedures are used in two main applicaitons: (a) exploratory analysis, where clusters
reveal homogeneous sub-groups within a large sample; (b) data
reduction, where high-dimensional item attribute vectors get reduced discrete
cluster labels \citep{jain1999data}.

@@

Several authors advocate choosing $k$ by performing a sequence of hypothesis
tests will null and alternative hypotheses of the form $H_0 : k = k_0$ and
$H_1: k > k_0$, starting with $k_0 = 1$ and proceeding sequentially with
higher values of $k_0$ until a test fails to reject $H_0$. The gap statistic
method typifies this class of methods, with a test statistic that measures the
within-cluster dispersion relative to what is expected under a reference
distribution \citep{tibshirani2001estimating}.
%Similar methods include those developed by \citet{TODO}.

Other authors have proposed choosing $k$ by using information
criteria.  For example, \citet{sugar2003finding} proposed an approach that
minimizes the estimated ``distortion'', the average distance per dimension.
Likewise, \citet{fraley2002model}'s model-based method fits Gaussian mixture model
models to the data, then selects the number of mixture components, $k$, using
the Bayesian Information Criterion (BIC).

@@

The procedure we propose in this report is based on a form of
cross-validation, and it is adaptive to the characteristics of the data
distribution. The essential idea is to devise a way to measure a form of
internal prediction error associated each choose of $k$, and then choose the
$k$ with the smallest associated error. We describe this method in detail in
Section~\ref{sec:meth}.  In Section~\ref{sec:self-consistent}, we prove that
our method is self-consistent. Then, in Section~\ref{sec:theory-gaussian}, we
analyze the performance of our method in the presence of Gaussian noise. The
theoretical analysis shows that the performance of our method degrades in the
presence of correlated noise; to fix this, we propose a correction for
correlated noise in Section~\ref{sec:corr-correct}. In
Sections~\ref{sec:simulations} and~\ref{sec:empirical-validation}, we
demonstrate that our method is competitive with other state-of-the-art
procedures in both simulated and real data sets. Then, in
Section~\ref{sec:application-yeast}, we apply our method to a Yeast cell cycle
dataset. We conclude with a short discussion in Section~\ref{sec:discussion}.

\section{Cross-validation for clustering}
\label{sec:meth}

\subsection{Problem statement}

Suppose that we are given a data matrix with $N$ rows and $P$ columns, and we
are tasked with choosing an appropriate number $k$ of clusters to use for
performing $k$-means clustering on the rows of the data matrix. Recall that
the $k$-means procedure takes a set of observations $\{ x_1, \dotsc ,x_n \}$
and finds a set of $k$ or cluster centers $A = \{ a_1, \dotsc, a_k \} $
minimizing the within cluster dispersion
\[
  W(A) = \sum_{i=1}^{n} \min_{a \in A} \|x_i - a\|^2.
\]
This implicitly defines a cluster assignment rule
\[
  g(x) = \argmin_{g \in \{1, \dotsc, k\}} \|x - a_g\|^2,
\]
with ties broken arbitrarily.
%We will assume that the $k$-means procedure
%finds an optimal choice of , $A$, but we will not assume that this solution is
%unique.

We can consider the problem of choosing $k$, the number of clusters, to be a
model selection problem. In other domains, especially supervised learning
problems like regression and classification, cross-validation is popular for
performing model selection.
In these settings, the data comes in the form of $N$
predictor-response pairs, $(X_1, Y_1), \dotsc, (X_N, Y_N)$, with $X_i \in
\R^{p}$ and $Y_i \in \R^{q}$.  The data can be represented as a matrix with
$N$ rows and $p + q$ columns.  We partition the data into $K$ hold-out
``test'' subsets, with $K$ typically chosen to be $5$ or $10$.  For each
``fold'' $r$ in the range $1, \dotsc, K$, we permute the rows of the data
matrix to get $\dataX$, a matrix with the $r$th test subset as its trailing
rows.  We partition $\dataX$ as
\[
  \dataX =
  \begin{bmatrix}
    \Xtrain & \Ytrain \\
    \Xtest  & \Ytest
  \end{bmatrix}.
\]
We use the training rows $[ \Xtrain\ \Ytrain ]$ to fit a regression model
$\hat Y = \hat Y(X)$, and then evaluate the performance of this model on the
test set, computing the cross-validation error $\|\Ytest - \hat Y(\Xtest)\|^2$
or some variant thereof.  We choose the model with the smallest
cross-validation error, averaged over all $K$ folds.

In unsupervised learning problems like factor analysis and clustering, the
features of the observations are not naturally partitioned into ``predictors''
and ``responses'', so we cannot directly apply the cross-validation procedure
described above.  For factor analysis, there are at least two versions of
cross-validation.  \citet{wold78cross} proposed a ``speckled'' holdout, where
in each fold we leave out a subset of the elements of the data matrix.  Wold's
procedure works well empirically, but does not have any theoretical support,
and it requires a factor analysis procedure that can handle missing data.
\citet{owen2009bi} proposed a scheme called ``bi-cross-validation'' wherein
each fold designates a subset of the data matrix columns to be response and a
subset of the rows to be test data.  This generalized a procedure due to
\citet{gabriel2002biblot}, who proposed holding out a single column and a
single row at each fold.

@@

In the sequel, we extend Gabriel cross-validation to the problem of selecting
the number of clusters, $k$, automatically, and we provide theoretical and
empirical support analogous to the consistency results proved by
\citet{owen2009bi}.

\subsection{Gabriel cross-validation}
\label{sec:gabriel-cv-algorithm}

Our version of Gabriel cross validation for clustering works by performing a
sequence of ``folds'' over the data. We use these folds to estimate a version
of prediction error (cross-validation error) for each possible value of $k$;
we then choose the value $\hat k$ with the smallest cross-validation error.

In each fold of our cross-validation procedure, we permute the rows and
columns of the data matrix and then partition the rows and columns as $N = n +
m$ and $P = p + q$ for positive integers $n$, $m$, $p$, and $q$.  We treat
the first $p$ columns as ``predictors'' and the last $q$ columns as
``responses''; similarly, we treat the first $n$ rows as ``training'' and the
last $m$ rows as ``test''.  In block form, the permuted data matrix is
\[
  \dataX
  =
  \begin{bmatrix}
    \Xtrain & \Ytrain \\
    \Xtest  & \Ytest
  \end{bmatrix},
\]
where
$\Xtrain \in \R^{n \times p}$,
$\Ytrain \in \R^{n \times q}$,
$\Xtest \in  \R^{m \times p}$,
and
$\Ytest \in  \R^{m \times q}$.

@@

TODO: describe theoretical support

\section{Self-consistency}
\label{sec:self-consistent}

An important property of any estimation procedure is that in the absence of of
noise, the procedure correctly estimates the truth. This property is called
``self-consistency'' \citep{tarpey96}. We will now show that Gabriel
cross-validation is self-consistent. That is, in the absence of noise, the
Gabriel cross-validation procedure finds the optimal number of clusters.

It will suffice to prove self-consistency for a single fold of the
cross-validation procedure.  As in in section~\ref{sec:gabriel-cv-algorithm}
we assume that the $P$ variables of the data set have been partitioned into
$p$ predictor variables represented in vector~$X$ and $q$ response variables
represented in vector~$Y$.  The $N$ observations have been divided into two
sets: $n$ train observations and $m$ test observations.  The following theorem
gives conditions for Gabriel CV to recover the true number of clusters in the
absence of noise.

@@

The result statement of Theorem~\ref{thm:self-consistency} follows from 
Lemmas~\ref{lem:self-consistency1} and~\ref{lem:self-consistency2}, which we
will now state and prove.

\begin{lemma}\label{lem:self-consistency1}
Suppose that the assumptions of Theorem~\ref{thm:self-consistency} are in
force.  If $k < K$, then $\CV(k) > 0$.
\end{lemma}
\begin{proof}
By definition,
\[
  \CV(k)
    =
      \sum_{i=n+1}^{n+m}
        \| Y_i - \bmuY (\hGX_i) \|^2,
\]
where $\bmuY(g)$ is the center of cluster $g$ returned from applying $k$-means
to $Y_1, \dotsc, Y_n$.  Assumptions~\ref{asn:self-consistency-noiseless}
and~\ref{asn:self-consistency-test}, imply that as $i$ ranges over the test
set $n+1, \dotsc, n+m$, the response $Y_i$ ranges over all distinct values in
$\{ \muY(1), \dotsc, \muY(K) \}$.
Assumption~\ref{asn:self-consistency-distinct-muy} implies that there are
exactly $K$ such distinct values.  However, there are only $k$ distinct values
of $\bmuY(g)$.  Thus, at least one summand
\(
  \| Y_i - \bmuY(\hGX_i) \|^2
\)
is nonzero.  Therefore,
\(
  \CV(k) > 0.
\)
\end{proof}

\begin{lemma}\label{lem:self-consistency2}
Suppose that the assumptions of Theorem~\ref{thm:self-consistency} are in
force.  If $k \geq K$, then $\CV(k) = 0$.
\end{lemma}
\begin{proof}
From assumptions~\ref{asn:self-consistency-noiseless},
\ref{asn:self-consistency-distinct-muy},
and~\ref{asn:self-consistency-train}, we know the cluster centers
gotten from applying $k$-means to $Y_1, \dotsc, Y_n$ must include
$\muY(1), \dotsc, \muY(K)$.  Without loss of generality, suppose that
$\bmuY(g) = \muY(g)$ for $g = 1, \dotsc, K$.  This implies that
$\hGY_i = G_i$ for $i = 1, \dotsc, n$.  Thus, employing
assumption~\ref{asn:self-consistency-noiseless} again, we get that
$\bmuX(g) = \muX(g)$ for $g = 1, \dotsc, K$.

@@

\section{Analysis under Gaussian noise}
\label{sec:theory-gaussian}

@@

Theorem~\ref{thm:self-consistency} tells us that the Gabriel cross-validation
method recovers the true number of clusters when the noise is negligible.
While this result gives us some assurance that the procedure is well-behaved,
we can bolster our confidence and gain insight into its workings by analyzing
its behavior in the presence of noise.  We first study the case of a single
cluster in two dimensions with correlated Gaussian noise.

@@

We confirm this result with a simulation.  We perform $10$ replicates.  In each
replicate, we generate $20000$ observations from a mean-zero bivariate normal
distribution with unit marginal variances and correlation $\rho$.  We perform
a single $2 \times 2$ fold of Gabriel cross-validation and report the
cross-validation mean squared error for the number of clusters $k$ ranging
from $1$ to $5$.  Figure~\ref{fig:nullcorr-equal} shows the cross-validation
errors for all $10$ replicates.  The simulation demonstrates that in the
Gabriel cross-validation criterion chooses the correct answer $k = 1$ whenever
$\rho < 0.5$; the criterion chooses $k \geq 2$ clusters whenever $|\rho| > 0.5$.

\begin{figure}[H]
\centering
\includegraphics[width=3in]{demo/nullcorr/equal.pdf}
\caption{Cross-validation error on $10$ replicates, with the number of
clusters $k$ ranging from $1$ to $5$.  Data is generated from two-dimensional
multivariate normal distribution with correlation $\rho$.  The Gabriel
cross-validation criterion chooses the correct answer $k = 1$ whenever
$|\rho| < 0.5$; the criterion chooses $k \geq 2$ clusters whenever $|\rho| > 0.5$.}
\label{fig:nullcorr-equal}
\end{figure}

Intuitively, when the correlation is high, the response feature, $Y$, looks
similar to the predictor feature, $X$. Prediction error on $X$ always
decreases with larger $k$. Thus, when the correlation is high, the prediction
error for $Y$ will also decrease with larger $k$. This explains why
cross-validation breaks down in the presense of strong correlation.

@@

\begin{proposition}
Suppose that $\{ (X_i, Y_i) \}_{i=1}^{n + m}$ is data from a single fold
of Gabriel cross-validation, where each $(X,Y)$ pair in $\R^{p+q}$ is an
independent draw from a mean-zero multivariate normal distribution with 
covariance matrix $\Sigma_{XY}= \left( \begin{smallmatrix} \Sigma_{XX} & \Sigma_{XY} \\ 
 \Sigma_{YX} & \Sigma_{YY} \end{smallmatrix}\right)$, with $\Sigma_{YY}$ has leading 
eigenvalue $\lambda_1$ and corresponding eigenvector $u_1$. In this case, the data are drawn
from a single cluster; the true number of clusters is~$1$.  If  $\frac{\sqrt{\lambda_1}}{2}
 > \frac{u^T_1\Sigma_{YX}\Sigma_{XY}u_1}{\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}}$,
then $\CV(1) < \CV(2)$ with probability tending to one as $m$ and $n$ increase.
\end{proposition}

\begin{proof}
Let $X$ and $Y$ be jointly multivariate normal distributed with mean $\mathbf{0}$ and covariance matrix $\Sigma_{XY}$, i.e.
\[	(X,Y) \sim \mathcal{N} \left( \mathbf{0}, \Sigma_{XY}\right)	\]
where $\Sigma_{XY}=\begin{bmatrix} \Sigma_{XX} & \Sigma_{XY} \\  \Sigma_{YX} & \Sigma_{YY} \end{bmatrix}$.

Let $\Sigma_{YY} = U \Lambda U^T$  be the eigendecomposition of $\Sigma_{YY}$, with leading eigenvalue $\lambda_1$ and corresponding eigenvector $u_1$. Then the centroid of $k$-means applying on $(y_1,..,y_n)$ is on the first PC 
of $Y$,\[	E(u^T_1 Y|u^T_1 Y>0) = \bar{\mu}^Y_1 =\sqrt{2 \lambda_1/\pi}u_1\] and 
\[	E(u^T_1 Y|u^T_1 Y<0) = \bar{\mu}^Y_2 =-\sqrt{2 \lambda_1/\pi}u_1\]
 where $u^T_1 Y \sim \mathcal{N}(0,\lambda_1)$.

To compute $\bar{\mu}^X_1 = E(X|u^T_1 Y>0)$, we need to know the conditional distribution $X|u^T_1 Y$. Since $(X,Y)$ has multivariate normal distribution, $(X,u^T_1 Y)$ also has a multivariate normal distribution with mean $\mathbf{0}$ and covariance matrix
$$\Sigma_{X,u^T_1 Y}=\begin{bmatrix} \Sigma_{XX} & \Sigma_{XY} u_1 \\  u^T_1 \Sigma_{YX} & \lambda_1 \end{bmatrix}$$
The conditional distribution $X|u^T_1 Y$ is hence normal with mean
 $$\mu_{X|u^T_1 Y} = \Sigma_{XY} u_1 \lambda^{-1}_1 u^T_1 Y $$
Therefore, 
\begin{align}
\bar{\mu}^X_1 &= E(X \mid u^T_1 Y>0) \nonumber \\ \nonumber
 			  &= E\left(E[X \mid u^T_1 Y] \mid u^T_1 Y>0\right) \\ \nonumber
 			  &=  E\left(\Sigma_{XY} u_1 \lambda^{-1}_1 u^T_1 Y \mid u^T_1 Y>0\right)\\ \nonumber
 			  &= \lambda^{-1}_1 \Sigma_{XY}u_1 E(u^T_1 Y \mid u^T_1 Y>0) \\ \nonumber
 			  &= \lambda^{-1}_1 \Sigma_{XY}u_1 \sqrt{2 \lambda_1/\pi} \\ \nonumber
 			  &= \sqrt{2 / \lambda_1 \pi} \Sigma_{XY}u_1
\end{align}
Similar calculation yields $\bar{\mu}^X_2 = -\sqrt{2 / \lambda_1 \pi} \Sigma_{XY}u_1$.
The decision rule to classify any observed value of $X$ to $\bar{\mu}^X_1$ is therefore
\[	(\bar{\mu}^X_1)^T X >0	\hspace{0.2in}\text{or} \hspace{0.2in} u^T_1\Sigma_{YX}X>0\] 
Since $u^T_1\Sigma_{YX}X$ is a linear combination of $X$, it also has normal distribution 
\[	\mathcal{N} \left( 0, u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1\right)	\]
And $(Y,u^T_1\Sigma_{YX}X)$ also have multivariate normal distribution with mean $\mathbf{0}$ 
and covariance matrix
\[
\begin{bmatrix}
\Sigma_{YY} & \Sigma_{YX}\Sigma_{XY}u_1  \\
u^T_1\Sigma_{YX}\Sigma_{XY} &  u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1
\end{bmatrix}
\]
The conditional distribution of $Y|u^T_1\Sigma_{YX}X$ is also multivariate normal with mean 
\[	
\mu_{Y|u^T_1\Sigma_{YX}X } = \Sigma_{YX}\Sigma_{XY}u_1 (u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1)^{-1}u^T_1\Sigma_{YX}X	
\]
The $Y$ center for $u^T_1\Sigma_{YX}X>0$ is
\begin{align}
\hat{\mu}^Y_1 &= E(Y|u^T_1\Sigma_{YX}X>0) \nonumber \\ \nonumber
 		      & =  \Sigma_{YX}\Sigma_{XY}u_1 (u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1)^{-1} E(u^T_1\Sigma_{YX}X \mid u^T_1\Sigma_{YX}X>0) \\ \nonumber
\end{align}
Note that $u^T_1\Sigma_{YX}X$ has normal distribution $\mathcal{N} \left( 0, u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1\right)$, so
\[
E(u^T_1\Sigma_{YX}X \mid u^T_1\Sigma_{YX}X>0) = \sqrt{2/\pi}\cdot\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}
\]
Therefore, we have the $Y$ center for $u^T_1\Sigma_{YX}X>0$ be
\begin{align*}
\hat{\mu}^Y_1 &= \sqrt{2/\pi}\cdot\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1} \hspace{0.1in} \Sigma_{YX}\Sigma_{XY}u_1 (u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1)^{-1} \\
&=\frac{\sqrt{2/\pi}}{\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}} \Sigma_{YX}\Sigma_{XY}u_1
\end{align*}

Recall that $\bar{\mu}^Y_1 =\sqrt{2 \lambda_1/\pi}u_1$, to judge if $CV(2) > CV(1)$, one only need to compare the distance between  $\hat{\mu}^Y_1$  and  $\bar{\mu}^Y_1$ with distance between  $\hat{\mu}^Y_1$ and grand mean $0$. By variance and bias decomposition of prediction MSE, when variance is the same, only bias influence the MSE.

After some linear algebra manipulation, we get
$||\hat{\mu}^Y_1 - \bar{\mu}^Y_1||^2 > ||\hat{\mu}^Y_1||^2$ or $CV(2) > CV(1)$ iff
\[
 \frac{\sqrt{\lambda_1}}{2} > \frac{u^T_1\Sigma_{YX}\Sigma_{XY}u_1}{\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}} 
\]
\end{proof}

@@

\section{Adjusting for correlation}
\label{sec:corr-correct}

When the correlation between dimensions are high, the proposed Gabriel CV method tends to overestimate the number $k$. A simply remedy for the high correlation is available if we assume common covariance structure among the $k$ clusters. 
The procedure consists of four steps as follows:
\begin{enumerate}
	\item Apply Gabriel CV method on the original data $\dataX$, get estimated number of cluster $\hat{k}$
	\item Estimate the pooled covariance matrix $\hat{\Sigma}$ from the $\hat{k}$ clusters.
	\item Let $\hat{\Sigma} = \Gamma\Lambda\Gamma'$ be the Eigendecomposition of $\hat{\Sigma}$, we rescale and rotate the original data $\dataX$ to get $ \widetilde{\dataX} = \dataX\Gamma\Lambda^{-1/2}Q$, where $Q$ is a random  orthonormal rotation matrix. 
	\item Apply Gabriel CV method again on the transformed data $ \widetilde{\dataX}$ to estimate $k$. 
\end{enumerate}
Such correlation-correction idea is based on the assumption that each observation $x_i$ in $\dataX$ can be expressed as
\[	x_i = \mu(G_i)+\varepsilon_i \]
where $\mu(G_i)$ is the true cluster center for $x_i$ and $\varepsilon_i$ is noise with covariance matrix $\Sigma$. To reduce correlation between dimensions, we would like to transform $\Sigma$ into a diagonal matrix. A simple way to achieve this goal is multiplying $\dataX$ by $\Sigma^{-1/2}$, so the transformed $x_i$

\[	\Sigma^{-1/2} x_i = \Sigma^{-1/2} \mu(G_i)+\Sigma^{-1/2} \varepsilon_i \]
has identity covariance matrix $\mathbf{I}$. This is exactly what step $3$ does. However, before we can transform the data, we need to estimate the covariance matrix $\Sigma$. Step $1$ and $2$ estimate the covariance matrix by first applying Gabriel CV method on $\dataX$ to get preliminary estimate $\hat{k}$ and then estimate $\Sigma$ through

\[	\hat{\Sigma} = \frac{1}{n-\hat{k}} \sum^n_{i=1} (x_i-\hat{\mu}_i)(x_i-\hat{\mu}_i)^T	\]
where $\hat{\mu}_i$ is the center of the cluster which contains $x_i$ when applying $k$-means on $\dataX$ with parameter $\hat{k}$.

In practice, $\hat{\Sigma}$ wouldn't be the efficient estimator of $\Sigma$ since step $1$ may yield $\hat{k} > k$. Nevertheless, above procedure works well with correlated noise as we shall see in the simulation section.

\section{Performance in simulations}
\label{sec:simulations}

@@

\section{Empirical validation}
\label{sec:empirical-validation}

@@

\section{Application on Yeast data set}
\label{sec:application-yeast}

The yeast Saccharomyces cerevisiae data set was collected by \cite{cho1998genome} to study the characterization of the cell cycle related processes of the budding yeast S. cerevisiae. To obtain data, cdc28 cells were synchronized by raising temperature to $37^\circ$C in late G1 phase, and then reinitiated by changing cells to cooler environment ($25^\circ$C). Data was collected at $17$ time points at every $10$ minutes interval covering almost $2$ complete cell cycles. Oligonucleotide microarrays was used to query $6,220$ gene expression profiles at these $17$ time points. Besides the original paper, several other studies have analyzed this data.

\cite{tavazoie1999systematic} applied $k$-means clustering on the preprocessed data with $k=30$ and discovered several periodic clusters. The preprocessing procedure includes choosing the most variable $2,945$ geness among the total $6,220$. The normalized dispersion across time points (s.d/mean) was used as the metric of variation for selection. Data at time points $90$ and $100$ minutes were removed because they were deemed unreliable. It results a $2,945$ by $15$ data matrix. The last step of the preprocessing procedure was to variance normalize each gene across the $15$ time points, which is done by subtracting the mean and then dividing by the standard deviation. The result is each gene has mean zero and standard deviation equal to one across the time points.

\cite{dortet2008model} performed model-based clustering on the preprocessed data of \cite{tavazoie1999systematic} and  found $27$ clusters. They classified the $27$ clusters into $3$ major types: cyclic gene clusters, stress gene clusters and the clusters neither cyclically nor sensitive to the induced stress.
